### 2022-11-08 NSF2022/Description/specificaim2.tex
working on SA2

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
\begin{itemize}
  \item Make clear that we know how to address NSI and fast summation
  \item Talk about collision as per talk with Bryce
  \item Wen's paper can be used to give numbers on how large or a
    problem could hypothetically be considered
  \item Emphasize that we've learned a lot from just $\mathcal(O)(100)$
    particles, but now it's time to do more
  \item Confinement and rheology
\end{itemize}

% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 2, we propose novel and efficient high-order numerical methods to
simulate dynamic assembly of particles under the HAP in a viscous
solvent. As described in~\eqref{eq:stokes} and illustrated in
Figure~\ref{fig:flow_map}, the solvent phase is modeled by the Stokes
equations for an incompressible, viscous fluid, and these equations are
coupled to the screened-Laplace equation~\eqref{eq:SL} through viscous
and hydrophobic stress balance.

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior problems for the order
parameter~\eqref{eq:SL} and the Stokes equations~\eqref{eq:stokes}. We
will develop methods based on combinations of volume and layer
potentials, and inexact Newton's method. Numerical methods to solve
these integral equations (IEs) have several advantages over their
PDE-based counterpart: by construction, they satisfy far-field and mass
conservation conditions; only $\partial\Omega$ needs to be discretized
for layer potentials; carefully chosen IEs result in a well-conditioned
linear system that is solved with a mesh-independent number of GMRES
iterations; and high-order or spectral accuracy is attainable with
appropriate quadrature methods. The PIs have demonstrated that IEs are a
powerful tool to simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo20}. Moreover, elements of these
formulations have been described in three dimensions~\cite{ying_2006,
manasthesis, rac-gre2016}. This proposal will extend the two-dimensional
methods to more general amphiphilic particle models that
satisfy~\eqref{eq:SL}, and all the methods can naturally be extended to
three dimensions. These methods will be motivated by models and analysis
developed in Specific Aims 1 and 3. \S\ref{subsec:AC} describes how the
non-linear Allen-Cahn equation~\eqref{eq:SL} will be iteratively solved
using an IE formulation. Then, \S\ref{subsec:NumericalIssues} describes
how we will combine preconditioning strategies with fast summation
methods to accelerate the time to solution, and
\S\ref{subsec:timeStepping} describes adaptive time stepping methods.

\subsubsection{Iterative methods for the Allen-Cahn HAP model}
\label{subsec:AC}

The PIs have developed numerical methods to solve~\eqref{eq:SL} when
$f(u) = u^2$---this choice results in the linear screened Laplace
equation. For a more general $f$, such as a double well potential, we
will interpret the solution of~\eqref{eq:SL} as the root finding problem
for
\begin{align}
  \label{eq:F}
  J[u] = -\rho^2 \Delta u + \tfrac{1}{2}f'(u).
\end{align}
This non-linear PDE will be solved with Newton's method
\begin{align*}
  u^{N+1} = u^{N} - \alpha (dJ^N)^{-1} J(u^N),
\end{align*}
where $dJ^N = -\rho^2 \Delta + \tfrac{1}{2}f''(u^N)$ is the Fr\'{e}chet derivative
of $F$, and $\alpha$ is the step size which will be determined using
backtracking. As expected, the biggest computational challenge is
solving $- \rho^2 \Delta v + \tfrac{1}{2}f''(u^{N}) v = J[u^N]$. Solving this
variable coefficient PDE with an integral equation is challenging since
a fundamental solution is not readily available. We propose to instead
use an inexact Newton's method where $dJ^{-1}$ is defined in terms of
the continuous, piecewise quadratic approximation
\begin{align*}
  f(u) \approx \tilde{f}(u) = bu + \begin{cases}
    a(u - u_0)^2, &\mbox{if } u \leq (u_0 + u_1)/2, \\
    (u - u_1)^2 + \tfrac{1}{4}(a-1)(u_0-u_1)^2 , &\mbox{if } u > (u_0 + u_1)/2,
  \end{cases}
\end{align*}
where $a>0$, and the resulting inexact Newton's method requires solving
the piecewise constant PDE
\begin{align}
  \label{eq:screenedPoisson}
  -\rho^2 \Delta v + \tfrac{1}{2}\tilde{f}''(u^{N})v = J[u^N],
\end{align}
which is a screened Poisson equation with piecewise a constant screening
coefficient. We test the efficacy of exact and inexact Newton's method
applied to~\eqref{eq:F} in $(0,1)$ by discretizing $dJ$ with a
second-order finite difference method. The choice of parameters $a$,
$b$, $u_0$, and $u_1$ result in the double well potential illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the max norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
piecewise constant approximate (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper in higher
dimensions.

@@

% ----------------------------------------------------------------------
\subsubsection{Accelerating the iterative solvers}
\label{subsec:NumericalIssues}

When solving for the density functions $\eta_0$, $\eta_1$, and $\sigma$,
a carefully chosen IE formulation only requires a mesh-independent
number of GMRES iterations~\cite{cam-ips-kel-mey-xue1996}. Therefore,
the required CPU time is proportional to the cost of a matrix-vector
multiplication that can be performed in optimal or near-optimal time
with the fast multipole method (FMM)~\cite{fmm5} and its
extensions~\cite{fmm1, fmm2, fmm3, fmm4, fmm6, fmm7, fmm8}. PI BQ is
experienced with applying FMMs for the Stokes
equation~\cite{qua-bir2014, bys-sha-qua2020} and the screened Laplace
equation~\cite{kro-qua2011, qua2011}. Furthermore, to reduce the number
of GMRES iterations, the efficacy of preconditioning strategies such as
the inverse Fast Multipole Method (IFMM)~\cite{cou-pou-dar2017} and
domain decomposition~\cite{che-bir2021} will be investigated. PI BQ used
the IFMM to precondition Stokes equations~\cite{qua-cou-dar2018}. A
suite of other preconditioners including sparse approximate
inverses~\cite{che2000} and multigrid~\cite{hem-sch1981, sch1982} will
be investigated.

% ----------------------------------------------------------------------
\subsubsection{Eliminating contact with adaptive time stepping}
\label{subsec:timeStepping}

>>>>

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
\begin{itemize}
%  \item Make clear that we know how to address NSI and fast summation
  \item Talk about collision as per talk with Bryce
%  \item Wen's paper can be used to give numbers on how large or a
%    problem could hypothetically be considered
%  \item Emphasize that we've learned a lot from just $\mathcal(O)(100)$
%    particles, but now it's time to do more
  \item Confinement and rheology
\end{itemize}

% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 2, we propose novel and efficient high-order numerical methods to
simulate dynamic assembly of particles under the HAP in a viscous
solvent. As described in~\eqref{eqn:stokes} and illustrated in
Figure~\ref{fig:flow_map}, the solvent phase is modeled by the Stokes
equations for an incompressible, viscous fluid, and these equations are
coupled to the non-linear screened-Laplace equation~\eqref{eqn:phase}
through viscous and hydrophobic stress balance.

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior problems for the order
parameter~\eqref{eqn:phase} and the Stokes equations~\eqref{eqn:stokes}.
We will develop methods based on combinations of volume and layer
potentials, and inexact Newton's method. As mentioned in
Section~\ref{sec:bie}, numerical methods to solve these integral
equations (IEs) have several advantages over their PDE-based
counterpart. The PIs have demonstrated that IEs are a powerful tool to
simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo20, fu-ryh-qua-you2022}. Moreover,
elements of these formulations have been described in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. This proposal
will extend the two-dimensional methods to more general amphiphilic
particle models and also to three dimensions. In addition, the methods
will allow us to use results we have shown for $\mathcal{O}(100)$
grains, and extend them to suspensions with 100X--1000X more grains.
These methods will be motivated by models and analysis developed in
Specific Aims 1 and 3. \S\ref{subsec:AC} describes how the non-linear
Allen-Cahn equation~\eqref{eq:SL} will be iteratively solved using an IE
formulation. Then, \S\ref{subsec:NumericalIssues} describes how we will
combine accurate quadrature and fast summation methods to accelerate the
time to accurate solutions. A natural extension to create simulations
that are realizable by experimentalists is to include confinement, and
this is discussed in \S\ref{subsec:Confinement}. This extension will
also allow us to study new rheological properties of the suspensions.
Finally, \S\ref{subsec:timeStepping} proposes collision-aware adaptive
time stepping methods.

% ----------------------------------------------------------------------
\subsubsection{Iterative methods for the Allen-Cahn HAP model}
\label{subsec:AC}

The PIs have developed numerical methods to solve~\eqref{eqn:phase} when
$f(u) = u^2$---this choice results in the linear screened Laplace
equation. For a more general $f$, such as a double well potential, we
will interpret the solution of~\eqref{eqn:phase} as the root finding
problem for
\begin{align}
  \label{eq:F}
  J[u] = -\rho^2 \Delta u + \tfrac{1}{2}f'(u).
\end{align}
This non-linear PDE will be solved with Newton's method
\begin{align*}
  u^{N+1} = u^{N} - \alpha (dJ^N)^{-1} J(u^N),
\end{align*}
where $dJ^N = -\rho^2 \Delta + \tfrac{1}{2}f''(u^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size which will be
determined using backtracking. As expected, the biggest computational
challenge is solving $- \rho^2 \Delta v + \tfrac{1}{2}f''(u^{N}) v =
J[u^N]$. Solving this variable coefficient PDE with an integral equation
is challenging since a fundamental solution is not readily available.
However, since $f''(u_0) = f''(u_1) = 2a(u_1 - u)0)^2 + 2c$, f' We
propose to instead use an inexact Newton's method where $dJ^{-1}$ is
defined in terms of the continuous, piecewise quadratic approximation
\begin{align*}
  f(u) \approx \tilde{f}(u) = bu + \begin{cases}
    a(u - u_0)^2, &\mbox{if } u \leq (u_0 + u_1)/2, \\
    (u - u_1)^2 + \tfrac{1}{4}(a-1)(u_0-u_1)^2 , &\mbox{if } u > (u_0 + u_1)/2,
  \end{cases}
\end{align*}
where $a>0$, and the resulting inexact Newton's method requires solving
the piecewise constant PDE
\begin{align}
  \label{eq:screenedPoisson}
  -\rho^2 \Delta v + \tfrac{1}{2}\tilde{f}''(u^{N})v = J[u^N],
\end{align}
which is a screened Poisson equation with piecewise a constant screening
coefficient. We test the efficacy of exact and inexact Newton's method
applied to~\eqref{eq:F} in $(0,1)$ by discretizing $dJ$ with a
second-order finite difference method. The choice of parameters $a$,
$b$, $u_0$, and $u_1$ result in the double well potential illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the max norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
piecewise constant approximate (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper in higher
dimensions.

@@

% ----------------------------------------------------------------------
\subsubsection{Increased accuracy and reduced CPU time}
\label{subsec:NumericalIssues}
While IE formulations of PDEs have several advantages, several
challenges must be overcome before the code can be used to simulate
systems with large numbers of grains. For example, when grains are close
to one another, which is inevitable in dense suspensions, the resulting
integrals are nearly-singular. This issue is well-known in the IE
community and several algorithms have been developed in the last 30
years. An incomplete list of methods that address near-singular
integration include panel-based methods~\cite{helsing_2008a,
helsing_tutorial_2012}, Quadrature by Expansions
(QBX)~\cite{din-hua-mar-tan2021, klo-bar-gre-one2013, ste-bar2022,
af2018sisc}, and interpolation-based methods~\cite{ying_2006,
qua-bir2014, rah-vee-zor-bir2015}. The proposed research, which is
described next,  will combine several methods including singularity
subtraction, analytic function theory, and spherical basis
representations.

In two-dimensions, analytic function theory can be used to derive a
quadrature method whose error is uniformly bounded~\cite{bar2014,
bar-wu-vee2015, ioa-pap-per1991}, but these works only show how to apply
the method to the two-dimensional Laplace and Stokes equations. PI BQ
recently showed that these quadrature methods can be extended to other
PDEs with singularity subtraction~\cite{che-lin-her-qua2021}. For
example, the fundamental solution of the screened Laplace
equation~\eqref{eqn:phase} involves the modified Bessel function of the
second kind, $K_0(\|\xx - \yy\|)$. The behavior of this function as $\yy
\rightarrow \xx$ is $K_0(\alpha \|\xx - \yy\|) = \log\|\xx - \yy\| +
S(\xx - \yy)$, where $S(\xx - \yy)$ is continuously differentiable. This
suggests a quadrature scheme for where the aforementioned quadrature
method is applied to the $\log\|\xx - \yy\|$ kernel, and high-order
methods~\cite{tre-wei2014} are applied to integrals involving the smooth
kernel $S(\xx - \yy)$.Similar techniques, which they call subtraction
techniques, have been considered by Camille Carvalho and her
colleagues~\cite{car2021, car2020, car-kha-kim2018}.  These methods are
particularly powerful when computing gradients of layer potentials which
is required, for example, when computing the water-order
stress~\eqref{eqn:stressbalance}.

In three dimensions, the quadrature methods are much more challenging.
Singularity subtraction methods can be applied, but analytic function
theory is no longer applicable. Singularity subtraction methods have
been applied using, for example, smoothed variations of
the~\cite{beale1, cor2001, tlu-bea2019}. Alternatively, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}. However, since we are interested in round (2D) and
spherical (3D) grains, we will exploit the fact that integrals can be
computed exactly after transforming the integrals to to a Fourier (2D)
or spherical harmonic (3D) basis. This technique can be used to evaluate
layer potentials for both Stokes~\cite{cor-vee2018} and the screened
Laplace equation~\cite{koh-cor-che-vee2021}.

Another facet to consider when choosing the quadrature rule is its
ability to be applied efficiently. Since carefully chosen IE
formulations require a mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}, the required CPU time is
proportional to the cost of a matrix-vector multiplication that can be
performed in optimal or near-optimal time with the fast multipole method
(FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3, fmm4, fmm6,
fmm7, ros-ols2016}. PI BQ is experienced with applying FMMs for the
Stokes equation~\cite{qua-bir2014, bys-sha-qua2020} and the screened
Laplace equation~\cite{kro-qua2011, qua2011}. Many of these methods rely
on analytic expansions of the PDE's fundamental solution, but recent
work has demonstrated that a kernel-independent method applied to
suspensions of 80,000 spheres can be simulated with only $\sim$10
seconds per time step on 1792 compute cores~\cite{Yan2019}.

%
%Furthermore, to reduce the
%number of GMRES iterations, the efficacy of preconditioning strategies
%such as the inverse Fast Multipole Method (IFMM)~\cite{cou-pou-dar2017}
%and domain decomposition~\cite{che-bir2021} will be investigated. PI BQ
%used the IFMM to precondition Stokes equations~\cite{qua-cou-dar2018}. A
%suite of other preconditioners including sparse approximate
%inverses~\cite{che2000} and multigrid~\cite{hem-sch1981, sch1982} will
%be investigated.

% ----------------------------------------------------------------------
\subsubsection{Incorporating confinement}
\label{subsec:Confinement}
In experiments, it is known that interactions between bounding walls and
grains cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modifications to the code will be performed so that confinement can be
considered. PI BQ has considered suspensions with confinement for
several projects including dynamic suspensions~\cite{qua-bir2014,
bys-sha-qua2020} and porous media~\cite{chi-moo-qua2020,
dea-qua-bir-jua2018, moo-che-chi-qua2022, qua-gan-you2021}.

% ----------------------------------------------------------------------
\subsubsection{Collision-aware suspensions with adaptive time stepping}
\label{subsec:timeStepping}
